---
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
header-includes:
  - \usepackage{setspace}\singlespacing
  - \raggedright
  - \pagenumbering{gobble}
mainfont: Source Sans Pro
bibliography:
  - "bib/refs.bib"
biblio-style: apa
csl: bib/apa.csl
---

```{r setup, include = FALSE}
needed_packages <- c("tidyverse", "here", "cmdstanr", "tidybayes", "ggdist",
                     "wjake", "showtext")
load_packages <- function(x) {
  if (!(x %in% installed.packages())) {
    install.packages(x, repos = "https://cran.rstudio.com/")
  }
  suppressPackageStartupMessages(require(x, character.only = TRUE))
}
vapply(needed_packages, load_packages, logical(1))

font_add(family = "Source Sans", regular = "SourceSansPro-Regular.ttf",
         italic = "SourceSansPro-Italic.ttf", bold = "SourceSansPro-Bold.ttf",
         bolditalic = "SourceSansPro-BoldItalic.ttf")
showtext_auto()
showtext_opts(dpi = 320)

options(knitr.table.format = "markdown")

knitr::opts_chunk$set(include = FALSE)
```

```{r functions}

```

```{r simulate-data}
set.seed(1213)


```

```{r fit-model}

```

```{r create-plot}

```

**Title:** Different Methods for Visualizing Uncertainty
\break

**Submitters:**

W. Jake Thompson, Ph.D.  
Accessible Teaching, Learning, and Assessment Systems (ATLAS); University of Kansas
\break

**Corresponding author:**

W. Jake Thompson, Ph.D.  
ATLAS, University of Kansas  
1122 West Campus Road  
Joseph R. Pearson Hall, Room 437  
Lawrence, KS 66045  
wjakethompson@gmail.com  

\newpage

**Title:** Different Methods for Visualizing Uncertainty
\break

**Caption:** A one sentence caption.
\break

```{r print-fig, include = TRUE, echo = FALSE, eval = FALSE}
#| out.width = "100%"

knitr::include_graphics(here("2023-uncertainty", "XXX.pdf"))
```

**Description:**

These uncertainty distributions are straightforward to creating using the {ggplot2} and {ggdist} packages in R [@R-ggplot2; @R-ggdist].
\break

**References**

